\begin{thebibliography}{99}
	\bibitem{thingie-paper} {
		D. Luposchainsky, A. Barato, H. Hinrichsen \\
		Phys. Rev. E 87, 042108 (2013) \\
		\emph{Strong fluctuation theorem for nonstationary nonequilibrium systems}
		}
	\bibitem{flow-paper} {
		D. Luposchainsky and H. Hinrichsen \\
		Journal of Statistical Physics (2013) \\
		DOI \texttt{10.1007/s10955-013-0863-0} \\
		\emph{Entropy production in continuous phase space systems}
		}
	\bibitem{seifert-review} {
		U. Seifert \\
		Rep. Prog. Phys. \textbf{75}, 126001 (2012) \\
		\emph{Stochastic thermodynamics, fluctuation theorems and molecular machines}
		}
	\bibitem{sf} {
		R.~Spinney and I.~Ford \\
		Phys. Rev. E \textbf{85}, 051113 (2012) \\
		\emph{Entropy production in full phase space for continuous stochastic dynamics}
		}
	\bibitem{sf-odd-even} {
		R.~Spinney and I.~Ford \\
		Phys. Rev. Lett. \textbf{108}, 170603 (2012) \\
		\emph{Nonequilibrium Thermodynamics of Stochastic Systems with Odd and Even Variables}
		}
	\bibitem{sde} {
		L. Evans \\
		Lecture notes on mathematics \\
		\emph{An Introduction to Stochastic Differential Equations}
		}
	\bibitem{sengupta} {
		A. Sengupta \\
		Summer term 2012 at Rutgers/USA \\
		\emph{Lecture notes on theoretical biophysics}
		}
	\bibitem{wissel} {
		C. Wissel \\
		Z. Phys. B \textbf{35}, 185 (1979) \\
		\emph{Manifolds of equivalent path integral solutions of the Fokker-Planck equation}
		}
	\bibitem{donoso} {
		J. Donoso, J. Salgado, M. Soler \\
		J. Phys. A: Math. Gen. \textbf{32} (1999) 3681-3695 \\
		\emph{Short-time propagators for nonlinear Fokker-Planck equations}
		}
	\bibitem{seifert-overdamped} {
		U. Seifert \\
		Phys. Rev. Lett. \textbf{95}, 040602 (2005)
		\emph{Entropy Production along a Stochastic Trajectory and an Integral Fluctuation Theorem}
	}
\end{thebibliography}

\TODO{sort in appearance order}
